### 2013-12-10 Publications/2013-ISBI14/02-introduction.tex
Fixing reviews
Final proposal and response letter draft

\section{INTRODUCTION}
\label{sec:intro}
In-vivo whole-brain connectivity analysis has been a
research topic of high interest for the last 5 years.
\Gls*{dmri} can be used to probe the
orientation of fiber bundles within the brain,
and is generally acquired with an \gls*{epi} sequence.
After signal reconstruction,
tractography algorithms draw a map of the sampled 
structures.
These maps can represent the actual trajectories
of fiber bundles (deterministic tractography) or
pixel-wise probability of connection to a certain origin
(probabilistic tractography). Finally, the
information about these connections is collected
into a network matrix that can be subjected to
the so-called ``connectome analysis''.

Among all the difficulties that such a complex workflow
raises \cite{jones_twenty-five_2010}, we will
address here the susceptibility-derived artifacts,
for which \Gls*{epi} schemes are especially sensitive.
Magnetic susceptibility disturbs the magnetic field
close to tissue interfaces. This inhomogeneity
of the field translates in a highly distorted
anatomy and a significant signal destruction in
certain regions of the brain 
(e.g. the orbitofrontal lobe, for the proximity of the
air surrounding sinuses). This artifact has been
well described, generally within the context of
functional MRI, which also uses \gls*{epi}.

Various approaches have been proposed to correct
for the distortions. \Gls*{fmb} methods
\cite{jezzard_correction_1995} rely on one extra
acquisition (\emph{field mapping}),
that probes the inhomogeneity of the B0 field.
A second theory-based breed of methodologies acquire a 
map of the \acrlong*{psf} of the \gls*{epi} readouts to correct 
for the artifact \cite{robson_measurement_1997}.
Another approach referred to as \gls*{reb}, acquires an
extra \gls*{epi} volume in the orthogonal or reversed
phase encoding that can be combined to remove the
geometric distortions
\cite{cordes_geometric_2000,chiou_simple_2000}.
The last family of methodologies acquires an
extra T2-weighted image, and uses it as an anatomical
reference to find the deformation map through nonlinear 
registration \cite{kybic_unwarping_2000,studholme_accurate_2000}.
These \gls*{t2b} methods usually map
the T2 image to the \emph{baseline} volume or \textit{b0}
of \gls*{dmri}. The choice of T2 is due to the strong
similarity of intensity distribution with respect the \textit{b0}.
More recent works report extensions or combined
approaches of existing techniques
\cite{andersson_how_2003,zaitsev_point_2004,%
holland_efficient_2010,andersson_comprehensive_2012}.

Even though the aforementioned techniques for distortion correction
have been studied \cite{zeng_image_2002,wu_comparison_2008},
the lack of a gold-standard limits benchmarking
strategies. Recently, Irfanoglu et al.
\cite{irfanoglu_effects_2012} raised the question
of distortion-derived impact on tractography.
In this work, we propose an evaluation framework
using a digital phantom designed for connectivity assessment.
This framework enabled us to compare several correction 
techniques and characterize their geometrical accuracy 
and the dropout compensation. Finally, we report their 
impact on subsequent tractography and resulting connectivity
matrices.

>>>>

\section{INTRODUCTION}
\label{sec:intro}
In-vivo whole-brain connectivity analysis has been a research topic of high
  interest for the last 5 yr.
\Gls*{dmri} can be used to probe the orientation of fiber bundles within the
  brain, and it is generally acquired with an \gls*{epi} sequence.
After signal reconstruction, tractography algorithms draw a map of the sampled
  structures.
These maps can represent the actual trajectories of fiber bundles
  (deterministic tractography) or pixel-wise probability of connection
  to a certain origin (probabilistic tractography).
Finally, the information about these connections is collected into a network
  matrix that is subjected to the ``connectome analysis''.

Among all the difficulties that such a complex workflow raises, % \cite{jones_twenty-five_2010},
  we will address here the susceptibility-derived artifacts,
  for which \Gls*{epi} schemes are especially sensitive.
Magnetic susceptibility disturbs the magnetic field close to tissue interfaces.
This inhomogeneity of the field translates into a highly distorted anatomy
  and a significant signal destruction in certain regions of the brain 
  (e.g. the orbitofrontal lobe, for the proximity of the air surrounding sinuses).
This artifact has been well described, generally within the context of functional MRI,
  which also uses \gls*{epi}.

Various approaches have been proposed to correct for this distortion.
\Gls*{fmb} methods \cite{jezzard_correction_1995} rely on one extra
  acquisition (\emph{field mapping}), that probes the inhomogeneity of the
  B0 field.
A second theory-based breed of methodologies acquire a map of the \acrlong*{psf}
  of the \gls*{epi} readouts to correct for the artifact \cite{robson_measurement_1997}.
Another approach referred to as \gls*{reb}, acquires an extra \gls*{epi} volume in the
  orthogonal or reversed phase encoding that can be combined to remove the geometric
  distortions \cite{cordes_geometric_2000,chiou_simple_2000}.
The last set of methodologies uses an extra T2-weighted image as anatomical reference
  to seek for the deformation map through nonlinear registration 
  \cite{kybic_unwarping_2000,studholme_accurate_2000}.
These \gls*{t2b} methods usually map the T2w image to the \emph{baseline} volume or
  \textit{b0} of \gls*{dmri}, as the latter is a T2 weighted image as well.
More recent works report extensions or combinations of existing techniques 
  \cite{andersson_how_2003,zaitsev_point_2004,%
  holland_efficient_2010,andersson_comprehensive_2012}.

Even though the aforementioned techniques for distortion correction have been studied
  \cite{zeng_image_2002,wu_comparison_2008}, the lack of a gold-standard limits benchmarking
  strategies.
Recently, Irfanoglu et al. \cite{irfanoglu_effects_2012} raised the question of distortion-derived
  impact on tractography.
In this work, we propose an evaluation framework using a digital phantom designed for connectivity
  assessment.
This framework enabled us to compare several correction techniques and characterize their geometrical
  accuracy and the dropout compensation.
Finally, we report their impact on subsequent tractography and connectivity.

### 2013-12-10 Publications/2013-ISBI14/03-methods.tex
Final proposal and response letter draft

\section{METHODS}

\subsection{Digital Phantom}
Based on the fiber geometries of the digital phantom 
created for the \emph{\Gls*{hardi} 
Reconstruction Challenge} held in ISBI 2013 
(San Francisco, US), we simulated high resolution 
(0.5mm isotropic) T1 (TE/TR=10/1500ms) and T2 
(TE/TR=90/5000ms) images, as well as two \gls*{dmri}
images (1.0mm isotropic, $b$=1200, 1 B0 image) with
32 and 64 evenly-distributed directions intended
for \gls*{dti} and \gls*{hardi} reconstructions, respectively.
Diffusion is modeled by a restricted and a hindered
compartment, similar to \cite{assaf_composite_2005}.
The phantom includes \gls*{wm} fiber bundles,
\gls*{gm} and \gls*{csf}. Physical properties
(T1/T2 times in sec) used in simulation are
($0.832\pm0.010$ / $79\times10^{-3}\pm0.6\times10^{-3}$)
for \gls*{wm},
($1.331\pm0.013$ / $110\times10^{-3}\pm2.0\times10^{-3}$) for
\gls*{gm} and ($3.5\pm0.1$ / $0.25\pm0.01$) for \gls*{csf}.

\subsection{Theory-based synthetic distortion}
\label{sec:distortion}
\Gls*{fmb} methodologies use a map
of the field in the scanner. More precisely, the
phase difference between two subsequent samplings
of the fieldmap. With that information, it is possible
to compute the theoretical displacement that each
voxel undergoes, the so-called \gls*{vsm}. The
most prominent feature of this \gls*{vsm} is that all
the shifts have the same orientation (parallel to the
phase-encoding direction of the \gls*{epi}) and their
magnitude and direction depend on the \gls*{epi} 
gradient increments (or \emph{blips}), and the actual
phase difference at the voxel.

In order to create a realistic distortion, we
generated a synthetic phase-difference map 
consistent with the phantom, using the 
tools distributed with the \emph{FSL} package 
\cite{jenkinson_fsl_2012}  and
standard parameters ($\Delta$TE=2.46~ms. for the
field mapping and \emph{effective dwell time} of 
0.77~ms. for the \gls*{epi}).
We defined two regions of smooth dephasing and 
computed the corresponding \gls*{vsm}. Amplitude
of the dephasing maps can be modulated, enabling
us to evaluate the impact of the distortion magnitude.
We generated several \glspl*{vsm} with increasing
maximum shifts (from 3.80 to 7.60~mm), covering 
the typical magnitudes of distortion observed 
in real datasets.

@@

In summary, we generated 
a full gold-standard containing realistic T1 and T2
at high resolution, \glspl*{dwi} acquired in two
different phase encoding directions, and a ground-truth
\gls*{dwi} data, which is not available for real datasets
(\autoref{fig:phantom}).

\begin{figure}[thpb]
   \centering
   \includegraphics[width=0.95\columnwidth]{Fig01-Phantom}
   \caption{Ground-truth digital phantom.
   A) T2-weighted image; B) undistorted \textit{b0} volume;
   C, D) distorted \textit{b0} volumes with opposed phase 
   encoding directions, maximum displacement of 3.80~mm.}
   \label{fig:phantom}
\end{figure}

\subsection{Correction methods}
\label{sec:correction}
Three out of four methods presented in \autoref{sec:intro}
were tested on the evaluation framework.

\Gls*{fmb} correction
is essentially the same as the one used for generating the 
distortion, on the inverse direction. Normally distributed 
noise was added to the fieldmap (for a signal-to-noise 
ratio of 20dB) before correction.

For \Gls*{reb} correction, we made use of the implementation found 
in \emph{FSL} (\texttt{topup})
that demands for the \textit{b0} of the reverse-encoding simulation.
In this second case, the \gls*{vsm} is inferred from the differences
between the two corresponding \textit{b0}.

Finally, for evaluating \Gls*{t2b},
we fine-tuned \emph{ANTs} (\url{http://stnava.github.io/ANTs}),
in order to register \textit{b0} to T2. To this end, we used a
multi-resolution scheme with 3 levels of subsampling and smoothing,
mutual information metric, and symmetric diffeomorphic transform 
(\texttt{SyN}). Several configurations of kernel widths for the 
regularization smoothers were tested, and finally selected 
0.5/1.0 voxels (gradient/deformation fields, respectively) 
for its best result. Additionally, undistorted images are 
corrected for dropout using the determinant
of the Jacobian of the deformation field.

@@

The framework supports two different methods for
\gls*{dwi} reconstruction and deterministic tractography.
Firstly, \emph{Diffusion Toolkit} \cite{wang_diffusion_2007}
for \gls*{dti}, is configured with 10 random
seeds per voxel by default. Secondly, \emph{MRTrix}
\cite{tournier_mrtrix:_2012} for \gls*{hardi}, with
default parameters set to use constrained spherical
deconvolution, maximum harmonic order of 6, and 150000
desired tracks.
For both options, the seeding regions can be set to use
either the distorted-corrected \gls*{wm} mask, or the 
regions used to generate the ground-truth. This second
seeding strategy mimics the usual procedure on real 
data, where regions are typically mapped from the
anatomically correct T1.

>>>>

\section{METHODS}\label{sec:methods}
\subsection{Digital Phantom}\label{sec:phantom}
Based on the fiber geometries of the digital phantom created
  for the \emph{\acrshort*{hardi} Reconstruction Challenge} held
  in ISBI 2013 (San Francisco, US), we simulated high resolution
  (0.5mm isotropic) T1 weighted (TE/TR= 10/1500ms) and T2w
  (TE/TR= 90/5000ms) images, as well as two \gls*{dmri} images
  (1.0mm isotropic, $b$=1200, 1 B0 image) with 32 and 64
  evenly-distributed directions intended for \gls*{dti} and 
  \gls*{hardi} reconstructions, respectively.
Diffusion is modeled by a restricted and a hindered compartment,
  similar to \cite{assaf_composite_2005}.
The phantom includes \gls*{wm} fiber bundles, \gls*{gm} and \gls*{csf}.
Physical properties (T1/T2 times in ms) used in simulation are
  ($832\pm10/79\pm0.6$) for \gls*{wm},
  ($1331\pm13/110\pm2.0$) for \gls*{gm} and
  ($3500\pm100$ / $250\pm10$) for \gls*{csf}.
T1w, T2w and \glspl*{dwi} were added normally distributed noise
  up to a \gls*{snr} of 30dB.

\subsection{Theory-based synthetic distortion}
\label{sec:distortion}
\Gls*{fmb} methodologies use a map of the field in the scanner.
More precisely, the phase difference between two subsequent samplings
  of the field map.
With that information, it is possible to compute the theoretical
  displacement that each voxel undergoes, the so-called \gls*{vsm}.
The most prominent feature of this \gls*{vsm} is that all the shifts
  have the same orientation (parallel to the phase-encoding direction
  of the \gls*{epi}) and their magnitude and direction depend on the 
  \gls*{epi} gradient increments (or \emph{blips})
  and the actual phase difference at the voxel.

In order to create a realistic distortion, we generated a synthetic and noise-free
  phase-difference map consistent with the phantom, using the tools distributed with
  the \emph{FSL} package %\cite{jenkinson_fsl_2012}
  (\url{www.fmrib.ox.ac.uk/fsl}) and standard parameters ($\Delta$TE=2.46~ms. for the
  field mapping and \emph{effective dwell time} of 0.77~ms for the \gls*{dwi}).
We defined two regions of smooth dephasing and computed the corresponding \gls*{vsm}.
Amplitude of the dephasing maps can be modulated, enabling us to evaluate the
  magnitude of distortion.
We generated several \glspl*{vsm} with increasing maximum shifts (from 3.80 to 7.60~mm),
  covering the typical magnitudes of distortion observed in real datasets.

@@

In summary, we generated a full gold-standard containing realistic T1w and T2w
  at high resolution, \glspl*{dwi} acquired in two different phase encoding directions,
  and a ground-truth \gls*{dwi} data, which is not available for real datasets
  (\autoref{fig:phantom}).
%The dataset is completed with two more images, the noise-free field map used for
%  generating the \gls*{vsm} and a field map with normally distributed noise for
%  a \gls*{snr} of 20dB.

\begin{figure}[thpb]
   \centering
   \includegraphics[width=0.95\columnwidth]{Fig01-Phantom}
   \caption{Ground-truth digital phantom.
   A) T2w; B) undistorted \textit{b0} volume;
   C, D) distorted \textit{b0} volumes with opposed phase 
   encoding directions, maximum displacement of 3.80~mm.}
   \label{fig:phantom}
\end{figure}

\subsection{Correction methods}
\label{sec:correction}
Three out of four methods presented in \autoref{sec:intro} were tested on the
  evaluation framework.
\Gls*{fmb} correction is the same as the one used for generating the distortion,
  working on the opposite direction.
Normally distributed noise was added to the synthetic field map (\gls*{snr}=20dB)
  before correction, for a more realistic evaluation.
\Gls*{reb} correction is implemented in \emph{FSL} (\texttt{topup}), and it demands for the \textit{b0}
  of the reverse-encoding simulation.
In this second case, the \gls*{vsm} is inferred from the differences between the two corresponding 
  \textit{b0}.
Finally, we included \Gls*{t2b} methods with \emph{ANTs} (\url{stnava.github.io/ANTs}),
  fine-tuned for \textit{b0}-to-T2w registration.
To this end, we used a multi-resolution scheme with 3 levels of subsampling and smoothing,
  mutual information metric, and the symmetric diffeomorphic transform model.
Several configurations of kernel widths for the regularization smoothers were tested, and finally
  selected 0.5/1.0 voxels (gradient/deformation fields, respectively) for its best result.
Additionally, undistorted images are corrected for dropout using the determinant of the Jacobian 
  of the deformation field.

@@

The framework supports two different methods for \gls*{dwi} reconstruction and deterministic
  tractography:
1) \emph{Diffusion Toolkit} (\url{trackvis.org/dtk}) %\cite{wang_diffusion_2007}
  for \gls*{dti}, is configured with 10 random seeds per voxel by default; and
2) \emph{MRTrix} %\cite{tournier_mrtrix:_2012}
  (\url{www.brain.org.au/software/mrtrix}) for \gls*{hardi}, with default parameters
  set to use constrained spherical deconvolution, maximum harmonic order of 6, and
  150000 desired tracks.
For both options, the seeding regions can be set to use either the distorted-corrected 
  \gls*{wm} mask, or the regions used to generate the ground-truth.
This second seeding strategy mimics the usual procedure on real data,
  where regions are typically mapped from the anatomically correct T1w.

### 2013-12-10 Publications/2013-ISBI14/04-results.tex
Final proposal and response letter draft

\section{RESULTS}
\label{sec:results}

All the modules described in the previous section were
conveniently integrated in workflows using \emph{nipype}
\cite{gorgolewski_nipype:_2011}. The choice of this tool
grants the reproducibility of the experiments,
and the evaluation workflow is publicly released.
The results of this study are summarized
in \autoref{fig:results} and \autoref{table:results}.
The remaining of this section provides extended descriptions
and interpretation of these findings.

\begin{figure*}[tpb]
   \centering
   \includegraphics[width=0.8\linewidth]{Fig02-Results}
   \caption{Visual assessment of the correction methods.}
   \label{fig:results}
\end{figure*}

\subsection{Geometrical correctness and signal recovery}

A summary of quantitative results, computed for the \gls*{dti} 
dataset with maximum shift of 3.80~mm,
is provided in \autoref{table:results}.
The best scores were obtained with the \gls*{reb} 
method, followed by \gls*{fmb}. 
The clear difference of accuracy between \gls*{reb} and \gls*{fmb}
with respect to \gls*{t2b} infers the latter may not be
an appropriate method for susceptibility correction. 
The described performance was constant for \gls*{fmb} 
and \gls*{reb} along the different magnitudes of 
distortion evaluated (maximum shifts of 3.80-7.60~mm).

\input{041-results-table}

\input{042-results-tractography}

A second workflow investigated the similarity of the recovered
signal with respect to the original (undistorted).
In this second study, \gls*{reb} performed
better than the other two methods, as reported
in \autoref{table:results}. \gls*{reb} scored a $91.0\%$
similarity index for the \textit{b0} volume and an average $97.65\%$
for the remaining \glspl*{dwi}. Again, the second
qualified was \gls*{fmb}, which achieved very close results
for the \glspl*{dwi} ($96.26\%$) but not as good for the \textit{b0}
volume. Visual inspection of the recovered data confirms the 
presented quantitative results (\autoref{fig:results}, first row).

\subsection{Impact on tractography and connectivity}

Connectivity matrices derived from the \gls*{dti} dataset
were completely hindered by the high complexity of the 
fibers contained
in the phantom. Tractography discovered successfully only 4 out 
of 27 existing connections from the original (undistorted) data,
with more than 45 false connections.
With the \gls*{hardi} dataset we found 23/27 connections,
but we still observed 40 false positives. Therefore, the 
experiments using the \gls*{dti} phantom were discarded. The
results presented in this subsection only refer to the 
\gls*{hardi} dataset.

Using the strongest distortion (maximum shift of 7.60~mm),
the number of detected connections
remained the same (23/27), slightly increasing the number of
false positives to 42. The immediate conclusion is that, with highly
complex anatomies (crossing, fanning, etc.) and limited number of
ground-truth connections, connectivity matrices are more sensitive 
to reconstruction and tractography than to distortions.

For the sake of completeness, \autoref{table:results-tractography}
reports the characteristic features of the tractography results
and connectivity matrices, for the fieldmap that caused a maximum
shift of 3.80~mm. A very similar outcome was obtained for
7.60~mm. These results, along with visual inspection
(\autoref{fig:results}, second and third rows), might point
to \gls*{fmb} as the best correction method.

\subsection{Discussion}

Even though all the surveyed methods produced visually 
sound results, our study suggested that \gls*{reb} is the 
most accurate method in terms of geometrical agreement and
signal dropout recovery. 
The \gls*{t2b} method did not achieve the necessary high-standards
to recommend its use. Nonetheless, we understand that specific
methods with anisotropic regularization that completely
restrict deformations to the phase-encoding direction would perform
significantly better than the standard method presented.
Geometrical correctness of \gls*{dwi} data is fundamental 
in connectivity analysis to spatially locate the
regions which will define the nodes of the final connectivity
matrix.

>>>>

\section{RESULTS}
\label{sec:results}
All the modules described in the previous section were
  conveniently integrated in workflows using \emph{nipype} % \cite{gorgolewski_nipype:_2011}
  (\url{nipy.sourceforge.net/nipype}).
The choice of this tool grants the reproducibility of the experiments,
  and the evaluation workflow is publicly released.
The results of this study are summarized in \autoref{fig:results} and
  \autoref{table:results}.
The remaining of this section provides extended descriptions and interpretation of these
  findings.
\begin{figure*}[tp]
   \centering
   \includegraphics[width=0.95\linewidth]{Fig02-Results}
   \caption{Visual assessment of the correction methods.}
   \label{fig:results}
\end{figure*}

\subsection{Geometrical correctness and signal recovery}\label{sec:results_geometry}
A summary of quantitative results, computed for the \gls*{dti} dataset with max. shift 3.80~mm,
  is provided in \autoref{table:results}.
The best scores were obtained with the \gls*{reb} method, followed by \gls*{fmb}.
The clear difference of accuracy between \gls*{reb} and \gls*{fmb} with respect to \gls*{t2b}
  infers the latter may not be an appropriate method for susceptibility correction.
The described performance was constant for \gls*{fmb} and \gls*{reb} along the different magnitudes 
  of distortion evaluated (max. shift 3.80-7.60~mm).
\input{041-results-table}
\input{042-results-tractography}
A second workflow investigated the similarity of the recovered signal with respect to the original
  (undistorted).
In this second study, \gls*{reb} performed better than the other two methods, as reported in
  \autoref{table:results}.
\Gls*{reb} scored a 91.0\% similarity index for the \textit{b0} volume and an average 97.65\%
  for the remaining \glspl*{dwi}.
Again, the second qualified was \gls*{fmb}, which achieved very close results for the \glspl*{dwi}
  (96.26\%) but not as good for the \textit{b0} volume.
Visual inspection of the recovered data confirms the presented quantitative results 
  (\autoref{fig:results}).

\subsection{Impact on tractography and connectivity}
Connectivity matrices derived from the \gls*{dti} dataset were completely hindered by the high complexity
  of the fibers contained in the phantom.
Tractography discovered successfully only 4 out of 27 existing connections from the original (undistorted) data,
  with more than 45 false connections.
With the \gls*{hardi} dataset we found 23/27 connections, but we still observed 40 false positives.
Therefore, the experiments using the \gls*{dti} phantom were discarded.
The results presented in this subsection only refer to the \gls*{hardi} dataset.

Using the strongest distortion (max. shift 7.60~mm), the number of detected connections
  remained the same (23/27), slightly increasing the number of false positives to 42.
The immediate conclusion is that, with highly complex anatomies (crossing, fanning, etc.)
  and limited number of ground-truth connections, connectivity matrices are more sensitive
  to reconstruction and tractography than to distortions.

For the sake of completeness, \autoref{table:results-tractography} reports the characteristic
  features of the tractography results and connectivity matrices, for the fieldmap that caused a max.
  shift of 3.80~mm.
A very similar outcome was obtained for 7.60~mm.
These results, along with visual inspection (\autoref{fig:results}), might point to \gls*{fmb} as
  the best correction method.

\subsection{Discussion}\label{sec:discussion}
Even though all the surveyed methods produced visually sound results, our study suggested that
  \gls*{reb} is the most accurate method in terms of geometrical agreement and signal dropout recovery.
The \gls*{t2b} method did not achieve the necessary high-standards to recommend its use.
Nonetheless, we understand that specific methods with anisotropic regularization that completely
  restrict deformations to the phase-encoding direction would perform significantly better than
  the standard method presented.
Geometrical correctness of \gls*{dwi} data is fundamental in connectivity analysis to spatially
  locate the regions which will define the nodes of the final connectivity matrix.
